Synthesising Recursive Functions for\\ First-Order Model Counting

### 2023-03-17 doc/talks/3_long/talk.tex
Ditto

\colorlet{redish}{color2!25}
\sethlcolor{redish}

@@

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{\today}
% TODO: update the date
% TODO: unify the use of colours
% TODO: do I want to mention forclift and/or crane?

@@

\begin{frame}{First-Order Knowledge Compilation: Before and After}
  \centering
  \begin{tikzpicture}
    \node (start) {\tiny $\begin{gathered}
        \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }\pP(x, y)\\
        \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\pP(x, y) \land \pP(x, z) \Rightarrow y=z\\
        \forall w, x \in \Gamma\text{. }\forall y \in \Delta\text{. }\pP(w, y) \land \pP(x, y) \Rightarrow w=x
      \end{gathered}$};
    \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of start] (kc) {Compilation};
    \node[below=0.09cm of kc] (graph) {
      \scalebox{.5}{
      \begin{forest}
        for tree={sn edges,grow=0,reversed,fill=red!50}
        [,ellipse,draw,name=gdr
        [,ellipse,draw
        [,ellipse,draw
        [,ellipse,draw
        [,ellipse,draw]
        [,ellipse,draw,name=ref]
        ]
        ]
        ]
        ]
        \draw[-Latex,bend left=20] (ref) to (gdr);
      \end{forest}
      }
    };
    \node[draw,rounded rectangle,fill=color2!50,below=0.1cm of graph] (interpretation) {Conversion};
    \node[below=0.3cm of interpretation] (initial) {\tiny $f(m, n) = \sum_{l=0}^m \binom{m}{l} [l<2] \times f(m-l, n-1)$};
    \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of initial] (simplification) {Simplification};
    \node[below=0.3cm of simplification] (simplified) {\tiny $f(m, n) = f(m, n-1) + mf(m-1, n-1)$};
    \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of simplified] (evaluation) {Evaluation};
    \node[below=0.3cm of evaluation] (end) {\includegraphics[height=0.4cm]{42.png}};
    \node[right=0.3cm of evaluation] (sizes) {\tiny $\begin{aligned}
        f(0,0) = 1&\text{{,} }f(m, 0) = 0\\
        m = 7&\text{{,} }n=2
      \end{aligned}$};
    %\node at (0.5, -0.55) {\textsc{Crane}};
    %\node[draw,fit={(kc) (interpretation)},inner ysep=6pt,yshift=5pt] {};
    \draw[-Latex,color=color2] (start) -- (kc);
    \draw[-Latex,color=color2] (kc) -- ($(graph)+(0,0.2)$);
    \draw[-Latex,color=color2] ($(graph)+(0,-0.3)$) -- (interpretation);
    \draw[-Latex,color=color2] (interpretation) -- (initial);
    \draw[-Latex,color=color2] (initial) -- (simplification);
    \draw[-Latex,color=color2] (simplification) -- (simplified);
    \draw[-Latex,color=color2] (simplified) -- (evaluation);
    \draw[-Latex,color=color2] (evaluation) -- (end);
    \draw[-Latex,color=color2] (sizes) -- (evaluation);
  \end{tikzpicture}
\end{frame}
% TODO: optimize edge lengths
% TODO: add forclift on the left, with its own example instance
% TODO: check if the math works out

% TODO: talk #4 might have some useful additions to this section

\begin{frame}{More Formally\ldots}
  \begin{definition}
    A \emph{first-order deterministic decomposable negation normal form computational graph} (FCG) is a
    \begin{itemize}
    \item directed graph
    \item (which is weakly connected)
    \item with a single source,
    \item labelled vertices,
    \item and ordered outgoing edges.
    \end{itemize}
  \end{definition}
\end{frame}

\begin{frame}{How to Interpret an FCG}
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50,fill on=<2>] (dr) {}
        child {node[fill=red!50,fill on=<3>] {}
                child {node[fill=red!50, fill on=<4>] {}
                  child {node[fill=red!50, fill on=<5>] {}
                    child {node[fill=red!50, fill on=<6>] {}}
                    child {node[fill=red!50, fill on=<7>] (ref) {}}
        }}};
        \draw[-latex, bend right] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.75\textwidth}
      \begin{align*}
        f(m, n) &= \onslide<3->{\sum_{l=0}^m \binom{m}{l}} \onslide<6->{[l<2]} \onslide<5->{\times} \onslide<7->{f(m-l, n-1)}\\
        \onslide<8>{&= f(m, n-1) + mf(m-1, n-1)}
      \end{align*}
      \onslide<6>{
        \[
          [\phi] =
          \begin{cases}
            1 & \text{if } \phi \\
            0 & \text{if } \neg\phi
          \end{cases}
          \]
      }
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}
  \frametitle<1>{New Rule 1: Generalised Domain Recursion}
  \frametitle<2>{New Rule 2: Constraint Removal}
  \frametitle<3>{New Rule 3: Identifying Possibilities for Recursion}
  \begin{columns}
    \begin{column}{0.15\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50, fill on=<1>] (dr) {}
        child {node {}
          child {node[fill=red!50, fill on=<2>] {}
            child {node {}
              child {node {}}
              child {node[fill=red!50, fill on=<3>] (ref) {}}
            }}};
        \draw[-latex, bend right,alt=<3>{red}{black}] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{overprint}
        \onslide<1>
          \begin{example}
            Input formula:
            \[
              \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
            \]
            Output formula (with a new constant \alert{$c \in \Gamma$}):
            \begin{gather*}
              \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(\alert{c}, y) \lor \neg \texttt{P}(\alert{c}, z) \\
              \begin{multlined}
                \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\alert{x \ne c} \land y \ne z \Rightarrow \\
                \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
              \end{multlined}
            \end{gather*}
            %% Output FCG:
            %% \begin{tikzpicture}[edge from parent/.style={draw,-latex}]
            %%   \node[draw,circle] {$\textsc{DR}$}
            %%   child {node {$\star$}}
            %%   ;
            %% \end{tikzpicture}
          \end{example}
        \onslide<2>
          \begin{example}
            Input formula (with a constant \alert{$c \in \Gamma$}):
            \begin{gather*}
              \begin{multlined}
                \forall x \in \alert{\Gamma}\text{. }\forall y, z \in \Delta\text{. }\alert{x \ne c} \land y \ne z \Rightarrow \\
                \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
              \end{multlined} \\
              \begin{multlined}
                \forall w,x \in \alert{\Gamma}\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \land \alert{x \ne c} \land w \ne x \Rightarrow \\
                \neg \texttt{P}(w, y) \lor \neg \texttt{P}(x, y)
              \end{multlined}
            \end{gather*}
            Output formula (with a new domain \alert{$\Gamma' \coloneqq \Gamma \setminus \{\, c \,\}$}):
            \begin{gather*}
              \forall x \in \alert{\Gamma'}\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z) \\
              \forall w,x \in \alert{\Gamma'}\text{. }\forall y \in \Delta\text{. }w \ne x \Rightarrow \neg \texttt{P}(w, y) \lor \neg \texttt{P}(x, y)
            \end{gather*}
          \end{example}
        \onslide<3>
          \begin{block}{Goal}
            Check if the input formula is isomorphic (up to domains) to a
            previously encountered formula.
          \end{block}
          \begin{block}{Rough Outline}
            \begin{enumerate}
              \item Consider pairs of `similar' clauses.
              \item Consider bijections between their sets of variables.
              \item Extend each such bijection to a map between sets of domains.
              \item If the bijection makes the clauses equal, and the domain map
                    is compatible with previous domain maps, move on to another
                    pair of clauses.
            \end{enumerate}
          \end{block}
      \end{overprint}
    \end{column}
  \end{columns}
\end{frame}
% We're still creating a vertex in a graph with one outgoing edge.

%% \begin{frame}{Compilation as Search}
%%   \begin{definition}
%%     A \alert{(search) state} is a tuple \structure{$(G, C, L)$}, where:
%%     \begin{itemize}
%%     \item \structure{$G$} is an FCG (or \texttt{null}),
%%     \item \structure{$C$} is a compilation cache that maps integers to sets of pairs $(\phi, v)$, where $\phi$ is a formula, and $v$ is a vertex of $G$ (which is used to identify opportunities for recursion),
%%     \item and \structure{$L$} is a list of formulas (that are yet to be compiled). (Note that the order is crucial!)
%%     \end{itemize}
%%   \end{definition}
%%   The search algorithm combines greedy and breadth-first search:
%% \end{frame}

@@

\begin{frame}{Summary \& Future Work}
  \begin{block}{Summary}
    The circuits hitherto used for FOMC become more powerful with:
    \begin{itemize}
    \item cycles,
    \item generalised domain recursion,
    \item and some more new compilation rules that support domain recursion.
    \end{itemize}
  \end{block}
  \begin{block}{Future Work}
    \begin{itemize}
    \item Automate:
      \begin{itemize}
      \item extracting and simplifying the definitions of functions,
      \item finding all base cases.
      \end{itemize}
    \item Open questions:
      \begin{itemize}
      \item What kind of \alert{sequences} are computable in this way?
      \item Would using a \alert{different logic} extend the capabilities of FOMC further?
      \end{itemize}
    \end{itemize}
  \end{block}
\end{frame}
% TODO: for future work, come back to the 'many ways to count' slide and add
% more arrows (e.g., beyond FOL)
\end{document}

>>>>

\colorlet{redish}{color2!25}
\sethlcolor{redish}
\setstcolor{black}

@@

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{\today}
% TODO: update the date
% TODO: unify the use of colours
% TODO: do I want to mention forclift and/or crane?
% TODO: maybe replace red!50 with one of the main colours
% TODO: \color -> \textcolor everywhere
% TODO: how do we capitalise blocks?

@@

\begin{frame}{First-Order Knowledge Compilation: Before and After}
  \begin{columns}[b]
    \begin{column}{0.4\textwidth}
      \centering
      \begin{tikzpicture}
        \node (start) {\tiny $\forall x \in \Delta\text{. } \pP(x) \lor \qQ(x)$};
        \node[draw,rounded rectangle,fill=color2!50,below=0.7cm of start] (kc) {Compilation};
        \node[below=0.7cm of kc] (graph) {
          \scalebox{.5}{
            \begin{forest}
              for tree={sn edges,fill=red!50,ellipse,draw}
              [[
              [
              []
              []
              ]
              [
              [
              []
              []
              ]
              []
              ]
              ]]
            \end{forest}
          }
        };
        \node[draw,rounded rectangle,fill=color2!50,below=0.7cm of graph] (evaluation) {Propagation};
        \node[below=0.7cm of evaluation] (end) {\includegraphics[height=0.4cm]{81.png}};
        \node[right=0.3cm of evaluation] (sizes) {\tiny $|\Delta| = 4$};
        \draw[-Latex,color=color2] (start) -- (kc);
        \draw[-Latex,color=color2] (kc) -- (graph);
        \draw[-Latex,color=color2] (graph) -- (evaluation);
        \draw[-Latex,color=color2] (evaluation) -- (end);
        \draw[-Latex,color=color2] (sizes) -- (evaluation);
      \end{tikzpicture}
    \end{column}
    \pause
    \begin{column}{0.6\textwidth}
      \centering
      \begin{tikzpicture}
        \node (start) {\tiny $\begin{gathered}
            \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }\pP(x, y)\\
            \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\pP(x, y) \land \pP(x, z) \Rightarrow y=z\\
            \forall w, x \in \Gamma\text{. }\forall y \in \Delta\text{. }\pP(w, y) \land \pP(x, y) \Rightarrow w=x
          \end{gathered}$};
        \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of start] (kc) {Compilation};
        \node[below=0.09cm of kc] (graph) {
          \scalebox{.5}{
            \begin{forest}
              for tree={sn edges,grow=0,reversed,fill=red!50}
              [,ellipse,draw,name=gdr
              [,ellipse,draw
              [,ellipse,draw
              [,ellipse,draw
              [,ellipse,draw]
              [,ellipse,draw,name=ref]
              ]
              ]
              ]
              ]
              \draw[-Latex,bend left=20] (ref) to (gdr);
            \end{forest}
          }
        };
        \node[draw,rounded rectangle,fill=color2!50,below=0.1cm of graph] (interpretation) {Conversion};
        \node[below=0.3cm of interpretation] (initial) {\tiny $f(m, n) = \sum_{l=0}^m \binom{m}{l} [l<2] \times f(m-l, n-1)$};
        \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of initial] (simplification) {Simplification};
        \node[below=0.3cm of simplification] (simplified) {\tiny $f(m, n) = f(m, n-1) + mf(m-1, n-1)$};
        \node[draw,rounded rectangle,fill=color2!50,below=0.3cm of simplified] (evaluation) {Evaluation};
        \node[below=0.3cm of evaluation] (end) {\includegraphics[height=0.4cm]{42.png}};
        \node[right=0.3cm of evaluation] (sizes) {\tiny $\begin{aligned}
                                                           f(0,0) = 1&\text{{,} }f(m, 0) = 0\\
                                                           m = 2&\text{{,} }n=7
                                                         \end{aligned}$};
        \draw[-Latex,color=color2] (start) -- (kc);
        \draw[-Latex,color=color2] (kc) -- ($(graph)+(0,0.2)$);
        \draw[-Latex,color=color2] ($(graph)+(0,-0.3)$) -- (interpretation);
        \draw[-Latex,color=color2] (interpretation) -- (initial);
        \draw[-Latex,color=color2] (initial) -- (simplification);
        \draw[-Latex,color=color2] (simplification) -- (simplified);
        \draw[-Latex,color=color2] (simplified) -- (evaluation);
        \draw[-Latex,color=color2] (evaluation) -- (end);
        \draw[-Latex,color=color2] (sizes) -- (evaluation);
      \end{tikzpicture}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Circuits vs Graphs}
  \begin{block}{Circuits
      \textcolor{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}\ldots}
    \begin{itemize}
      \item \ldots extend d-DNNF circuits
            \textcolor{gray}{\parencite{DBLP:journals/jancl/Darwiche01}} for
            propositional knowledge compilation with \alert{more node types}
      \item \ldots are \alert{acyclic}.
    \end{itemize}
  \end{block}
  \pause
  \begin{block}{First-Order Computational Graphs (FCGs) are\ldots}
    directed \alert{\st{acyclic}} (weakly connected) graphs with:
    \begin{itemize}
    \item a single source,
    \item labelled nodes,
    \item and ordered outgoing edges.
    \end{itemize}
  \end{block}
\end{frame}

% TODO: add arrows that explain the meaning/definition of each node type
\begin{frame}{How to Interpret an FCG}
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=12mm,level distance=15mm]
        \node[fill=red!50,fill on=<2>] (dr) {\textsc{GDR}}
        child {node[fill=red!50,fill on=<3>] {$\bigvee$}
                child {node[fill=red!50, fill on=<4>] {\textsc{CR}}
                  child {node[fill=red!50, fill on=<5>] {$\land$}
                    child {node[fill=red!50, fill on=<6>] {$\bot$}}
                    child {node[fill=red!50, fill on=<7>] (ref) {\textsc{Ref}}}
        }}};
        \draw[-latex, bend right] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.75\textwidth}
      \begin{align*}
        f(m, n) &= \onslide<3->{\sum_{l=0}^m \binom{m}{l}} \onslide<6->{[l<2]} \onslide<5->{\times} \onslide<7->{f(m-l, n-1)}\\
        \onslide<8>{&= f(m, n-1) + mf(m-1, n-1)}
      \end{align*}
      \onslide<6>{
        \[
          [\phi] =
          \begin{cases}
            1 & \text{if } \phi \\
            0 & \text{if } \neg\phi
          \end{cases}
          \]
      }
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}
  \frametitle<1>{New Rule 1: Generalised Domain Recursion}
  \frametitle<2>{New Rule 2: Constraint Removal}
  \frametitle<3>{New Rule 3: Identifying Possibilities for Recursion}
  \begin{columns}
    \begin{column}{0.15\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50, fill on=<1>] (dr) {}
        child {node {}
          child {node[fill=red!50, fill on=<2>] {}
            child {node {}
              child {node {}}
              child {node[fill=red!50, fill on=<3>] (ref) {}}
            }}};
        \draw[-latex, bend right,alt=<3>{red}{black}] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{overprint}
        \onslide<1>
          \begin{example}
            Input formula:
            \[
              \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
            \]
            Output formula (with a new constant \alert{$c \in \Gamma$}):
            \begin{gather*}
              \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(\alert{c}, y) \lor \neg \texttt{P}(\alert{c}, z) \\
              \begin{multlined}
                \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\alert{x \ne c} \land y \ne z \Rightarrow \\
                \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
              \end{multlined}
            \end{gather*}
            %% Output FCG:
            %% \begin{tikzpicture}[edge from parent/.style={draw,-latex}]
            %%   \node[draw,circle] {$\textsc{DR}$}
            %%   child {node {$\star$}}
            %%   ;
            %% \end{tikzpicture}
          \end{example}
        \onslide<2>
          \begin{example}
            Input formula (with a constant \alert{$c \in \Gamma$}):
            \begin{gather*}
              \begin{multlined}
                \forall x \in \alert{\Gamma}\text{. }\forall y, z \in \Delta\text{. }\alert{x \ne c} \land y \ne z \Rightarrow \\
                \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
              \end{multlined} \\
              \begin{multlined}
                \forall w,x \in \alert{\Gamma}\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \land \alert{x \ne c} \land w \ne x \Rightarrow \\
                \neg \texttt{P}(w, y) \lor \neg \texttt{P}(x, y)
              \end{multlined}
            \end{gather*}
            Output formula (with a new domain \alert{$\Gamma' \coloneqq \Gamma \setminus \{\, c \,\}$}):
            \begin{gather*}
              \forall x \in \alert{\Gamma'}\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z) \\
              \forall w,x \in \alert{\Gamma'}\text{. }\forall y \in \Delta\text{. }w \ne x \Rightarrow \neg \texttt{P}(w, y) \lor \neg \texttt{P}(x, y)
            \end{gather*}
          \end{example}
        \onslide<3>
          \begin{block}{Goal}
            Check if the input formula is isomorphic (up to domains) to a
            previously encountered formula.
          \end{block}
          \begin{block}{Rough Outline}
            \begin{enumerate}
              \item Consider pairs of `similar' clauses.
              \item Consider bijections between their sets of variables.
              \item Extend each such bijection to a map between sets of domains.
              \item If the bijection makes the clauses equal, and the domain map
                    is compatible with previous domain maps, move on to another
                    pair of clauses.
            \end{enumerate}
          \end{block}
      \end{overprint}
    \end{column}
  \end{columns}
\end{frame}

%% \begin{frame}{Compilation as Search}
%%   \begin{definition}
%%     A \alert{(search) state} is a tuple \structure{$(G, C, L)$}, where:
%%     \begin{itemize}
%%     \item \structure{$G$} is an FCG (or \texttt{null}),
%%     \item \structure{$C$} is a compilation cache that maps integers to sets of pairs $(\phi, v)$, where $\phi$ is a formula, and $v$ is a node of $G$ (which is used to identify opportunities for recursion),
%%     \item and \structure{$L$} is a list of formulas (that are yet to be compiled). (Note that the order is crucial!)
%%     \end{itemize}
%%   \end{definition}
%%   The search algorithm combines greedy and breadth-first search:
%% \end{frame}

@@

\begin{frame}{Summary \& Future Work}
  \begin{block}{Summary}
    The circuits hitherto used for FOMC become more powerful with:
    \begin{itemize}
    \item cycles,
    \item generalised domain recursion,
    \item and some more new compilation rules that support domain recursion.
    \end{itemize}
  \end{block}
  \begin{block}{Future Work}
    \begin{itemize}
    \item Automate:
      \begin{itemize}
      \item extracting and simplifying the definitions of functions,
      \item finding all base cases.
      \end{itemize}
    \item Open questions:
      \begin{itemize}
      \item What kind of \alert{sequences} are computable in this way?
      \item Would using a \alert{different logic} extend the capabilities of FOMC further?
      \end{itemize}
    \end{itemize}
  \end{block}
\end{frame}
% TODO: for future work, come back to the 'many ways to count' slide and add
% more arrows (e.g., beyond FOL)
% TODO: also incorporate the last slide of the 4th talk
\end{document}
